A theorical-practical approach to manifold learning

### 2018-11-05 DimReduction/Mds/Documentation/Article_mds.tex
Adding implementation section to Mds documentation

\begin{abstract}
    This work presents an theorical approximation to the problem and provide an
    algorithmic solution easy and understandable. The approach to the problem is
    reconsider by the construction of the matrix $D^x$ with KNN. Creating the
    possibility to improve the results using other ways to calculate the distance
    matrix
\end{abstract}

@@

@@

    \subsection{Points from distances}

    The set up of the problem is a matrix of euclidean distances $D$, where each
    entry is the square of the distance between two points in a $d$ dimensional
    space.

@@

    The unknown of the problem is a matrix $X$ that cotains the position of the
    $n$ points in the $d$ dimensional space.

@@

    But there is a restriction to make sure that the matrix is centered at 0.

@@

    \subsection{Solution}

    First rewrite each entry of matrix $D$ in a vector form

@@

    Then expand distance $d_{ij}^2$

@@

    The goal is to express all the dot products of $x$ vectors in terms of the
    $d_{ij}^2$ distances. Consider the following expression

@@

    Remember that

@@

    Finally, the first approximation to the problem can be constructed

@@

    Notice that

    \begin{equation*}
        \sum_{i=1}^n x_i^t x_i = \sum_{j=1}^n x_j^t x_j
    \end{equation*}

    So

    \begin{equation*}
        \begin{aligned}
            x_i^t x_j&=
            \frac{1}{2} \left[ \frac{1}{n} \sum_{j=1}^n d_{ij}^2 - d_{ij}^2 + \frac{1}{n} \sum_{i=1}^n d_{ij}^2 - \frac{2}{n} \sum_{i=1}^n x_i^t x_i \right]\\
        \end{aligned}
    \end{equation*}

@@

    Now the expression can be completed
    \begin{equation*}
        \begin{aligned}
            x_i^t x_j
            &=
            \frac{1}{2} \left[ \frac{1}{n} \sum_{i=1}^n d_{ij}^2 - d_{ij}^2 + \frac{1}{n} \sum_{j=1}^n d_{ij}^2 - \frac{1}{n^2}\sum_{i=1}^n \sum_{j=1}^n d_{ij}^2 \right]\\
        \end{aligned}
    \end{equation*}

    This computation can be expressed in matrix form to simplify the formula.
    First define

@@

    Now a matrix $B$ can be defined as

@@

    This formula gives a way to compute all the dot products between the points
    that makes up the matrix $X$. Because of this the matrix $B$ can also
    be expressed as

@@

    Which gives a way to compute all the points of the matrix $X$.

    \section{Mds optimization problem}

    if $d_{ij}^{(x)}$ es euclidean, then $d_{ij}^{(x)} = x_i^t x_j$, and hence

    or in matrix form

    As shown in section \ref{norm_trace}, the norm of a matrix can be transformed
    to a trace, so the problem can be redefined as

    \begin{equation*}
        \begin{aligned}
            A = X^tX &- Y^tY\\
            \underset{Y}{\text{min}} \quad \lVert X^t X - Y^t Y \rVert^2
            &=
            \underset{Y}{\text{min}} \quad \lVert A \rVert^2\\
            &=
            \underset{Y}{\text{min}} \quad Tr(A^tA) \\
        \end{aligned}
    \end{equation*}\\

    Now

    Which means that $A^tA = AA = A^2$, and hence

    \begin{equation*}
        \begin{aligned}
            \underset{Y}{\text{min}} \quad Tr(A^tA)
            &=
            \underset{Y}{\text{min}} \quad Tr(A^2)\\
            &=
            \underset{Y}{\text{min}} \quad Tr[(X^tX - Y^tY)^2]
        \end{aligned}
    \end{equation*}\\

    now as shown in section \ref{spectral_decomp}, $X^tX$ and $Y^tY$ can be
    expresed as

    Then

    We know that $V^tV = I = VV^t$, then using the proof presented in section \ref{circular_trace}

    \begin{equation*}
        \begin{aligned}
            A =& V\Lambda V^t - Q \hat{\Lambda} Q^t\\
            Tr[(V^tAV)^2] &= Tr[V^tAVV^tAV]\\
                          &= Tr[V^tAIAV]\\
                          &= Tr[AIAVV^t]\\
                          &= Tr[AIAI]\\
                          &= Tr[AA]\\
                          &= Tr[A^2]\\
        \end{aligned}
    \end{equation*}\\

    \begin{equation*}
        \begin{aligned}
            Tr[(V^tAV)^2] &= Tr[A^2]\\
            Tr[(V^t(V\Lambda V^t - Q \hat{\Lambda} Q^t)V)^2] &= Tr[(V\Lambda V^t - Q \hat{\Lambda} Q^t)^2]\\
            Tr[(V^tV\Lambda V^tV - V^tQ \hat{\Lambda} Q^tV)^2] &=\\
            Tr[(I\Lambda I - V^tQ \hat{\Lambda} Q^tV)^2] &=\\
            Tr[(\Lambda - V^tQ \hat{\Lambda} Q^tV)^2] &=\\
        \end{aligned}
    \end{equation*}

    Now

    \begin{equation*}
        \begin{aligned}
            G &= V^tQ\\
            Tr[(\Lambda - V^tQ \hat{\Lambda} Q^tV)^2] &= Tr[(\Lambda - G \hat{\Lambda} G^t)^2]\\
        \end{aligned}
    \end{equation*}\\

    Which means that the problem can be redefined one more time

    \begin{equation*}
        \begin{aligned}
        \underset{Q,\hat{\Lambda}}{\text{min}} \quad Tr[(V\Lambda V^t - Q \hat{\Lambda} Q^t)^2]
        &=
        \underset{G,\hat{\Lambda}}{\text{min}} \quad Tr[(\Lambda - G \hat{\Lambda} G^t)^2]
        \end{aligned}
    \end{equation*}\\

    Now

    \begin{equation*}
        \begin{aligned}
        Tr[(\Lambda - G \hat{\Lambda} G^t)^2]
        &=
        Tr[\Lambda^2 - 2 \Lambda G \hat{\Lambda} G^t + (G \hat{\Lambda} G^t)^2]\\
        &=
        Tr[\Lambda^2 - 2 \Lambda G \hat{\Lambda} G^t + G \hat{\Lambda} G^t G \hat{\Lambda} G^t]\\
        \end{aligned}
    \end{equation*}

    But (Knowing that $Q^tQ = I = QQ^t$)

    \begin{equation*}
        \begin{aligned}
            G^tG &= (V^tQ)^t V^tQ\\
                 &= (Q^tV) V^tQ\\
                 &= Q^tQ\\
                 &= I\\ \\
            GG^t &= V^tQ (V^tQ)^t\\
                 &= V^tQ (Q^tV)\\
                 &= V^tV\\
                 &= I\\ \\
        \end{aligned}
    \end{equation*}

    \subsection{Dark part}

    We know that $\frac{\delta Tr[\Lambda G \hat{\Lambda} G^t]}{\delta G} = 2\Lambda G \hat{\Lambda}$\\\\
    We also know that $\frac{\delta Tr(\Lambda^2)}{\delta G} = 0 = \frac{\delta Tr(\hat{\Lambda}^2)}{\delta G}$\\\\
    Finally we know that the answer for a fixed $\hat{\Lambda}$ is $G=I$\\ \\

    Now, coming back to the original problem

    So we can take the first $p$ eigenvalues of $\Lambda$ to define $\hat{\Lambda}$,
    and finally

    which implies that $Y = \hat{\Lambda}^{1/2} V^t$

    \section{Miscellaneous proofs}

@@

    Concider a matrix B that is computed as $A^tA$

@@

    Then an element of B can be computed as

@@

>>>>

\begin{abstract}
    This work presents an theorical and practical approximation to the problem
    of non-linear dimensionality reduction using the multidimensional scaling
    method and provide an algorithmic implementation easy and understandable. In
    this approach a distance matrix $D^x$ is constructed to approximate the
    distance over a manifold by solving shortest path problems over a graph that
    represents a multidimensional structure. This matrix is used as input to the
    method, which is tested with 3 dimensional manifolds. This technique gives a
    way to improve the results using other ways to calculate the distance matrix.
\end{abstract}

@@

Dimensionality reduction is a technique

\cite{manifold}

\cite{dimension}

\cite{mds}

@@

    As mentioned before, one of the goals of this work is to create a theorical
    development of the multidimentional scaling method. This section shows a
    detail explanation of the therical aspects of the method taking \cite{proof}
    as a base point.

    \vspace{0.5cm}

    The set up of the problem is a distance matrix $D^x$, where each
    entry is the square of the distance between two points in a $d$ dimensional
    space. The goal of this part is to find a distance matrix in a low
    dimensional space that is as similar as possible to the multidimensional one.
    This means that the problem can be expressed in mathematical terms as the
    following optimization problem

    Since each entry of the matrix $D^x$ can be expressed as
    $d_{ij}^{(x)} = x_i^t x_j$, the problem can be rewritten

    For simplicity it can also be expressed in terms of matrix opperations

    In section \ref{norm_trace}, it was shown that the norm of a matrix can be
    transformed to a trace, so the problem can be redefined once again

    \begin{equation*}
        \begin{aligned}
            A = X^tX &- Y^tY\\
            \underset{Y}{\text{min}} \quad \lVert X^t X - Y^t Y \rVert^2
            &=
            \underset{Y}{\text{min}} \quad \lVert A \rVert^2\\
            &=
            \underset{Y}{\text{min}} \quad Tr(A^tA) \\
        \end{aligned}
    \end{equation*}

    Now, if the transpose of the matrix $A$ is considered, the result would be

    which means that $A^tA = AA = A^2$, and hence

    \begin{equation*}
        \begin{aligned}
            \underset{Y}{\text{min}} \quad Tr(A^tA)
            &=
            \underset{Y}{\text{min}} \quad Tr(A^2)\\
            &=
            \underset{Y}{\text{min}} \quad Tr[(X^tX - Y^tY)^2]
        \end{aligned}
    \end{equation*}

    In section \ref{spectral_decomp}, it was shown that $X^tX$ and $Y^tY$ can be
    expresed as

    Then, the problem can be transformed

    Now, since $V^tV = I = VV^t$, then using the proof presented in section
    \ref{circular_trace} the following propertie can be deducted

    \begin{equation*}
        \begin{aligned}
            A =& V\Lambda V^t - Q \hat{\Lambda} Q^t\\
            Tr[(V^tAV)^2] &= Tr[V^tAVV^tAV]\\
                          &= Tr[V^tAIAV]\\
                          &= Tr[AIAVV^t]\\
                          &= Tr[AIAI]\\
                          &= Tr[AA]\\
                          &= Tr[A^2]\\
        \end{aligned}
    \end{equation*}

    \begin{equation*}
            \hspace{-1cm}
        \begin{aligned}
            Tr[(V^tAV)^2] &= Tr[A^2]\\
            Tr[(V^t(V\Lambda V^t - Q \hat{\Lambda} Q^t)V)^2] &= Tr[(V\Lambda V^t - Q \hat{\Lambda} Q^t)^2]\\
            Tr[(V^tV\Lambda V^tV - V^tQ \hat{\Lambda} Q^tV)^2] &=\\
            Tr[(I\Lambda I - V^tQ \hat{\Lambda} Q^tV)^2] &=\\
            Tr[(\Lambda - V^tQ \hat{\Lambda} Q^tV)^2] &=\\
        \end{aligned}
    \end{equation*}

    At this point, a new matrix $G$ can be defined to simplify the expression

    \begin{equation*}
        \begin{aligned}
            G &= V^tQ\\
            Tr[(\Lambda - V^tQ \hat{\Lambda} Q^tV)^2] &= Tr[(\Lambda - G \hat{\Lambda} G^t)^2]\\
        \end{aligned}
    \end{equation*}

    Hence the problem can be redefined one more time

    \begin{equation*}
        \hspace{-0.5cm}
        \begin{aligned}
        \underset{Q,\hat{\Lambda}}{\text{min}} \quad Tr[(V\Lambda V^t - Q \hat{\Lambda} Q^t)^2]
        &=
        \underset{G,\hat{\Lambda}}{\text{min}} \quad Tr[(\Lambda - G \hat{\Lambda} G^t)^2]
        \end{aligned}
    \end{equation*}

    Now, using that knowledge, the main expression can be expanded

    \begin{equation*}
        \begin{aligned}
        Tr[(\Lambda - G \hat{\Lambda} G^t)^2]
        &=
        Tr[\Lambda^2 - 2 \Lambda G \hat{\Lambda} G^t + (G \hat{\Lambda} G^t)^2]\\
        =
        &Tr[\Lambda^2 - 2 \Lambda G \hat{\Lambda} G^t + G \hat{\Lambda} G^t G \hat{\Lambda} G^t]\\
        \end{aligned}
    \end{equation*}

    But, knowing that $Q^tQ = I = QQ^t$, the following properties can be developed

    \begin{equation*}
        \begin{aligned}
            G^tG &= (V^tQ)^t V^tQ\\
                 &= (Q^tV) V^tQ\\
                 &= Q^tQ\\
                 &= I
        \end{aligned}
    \end{equation*}

    \begin{equation*}
        \begin{aligned}
            GG^t &= V^tQ (V^tQ)^t\\
                 &= V^tQ (Q^tV)\\
                 &= V^tV\\
                 &= I\\ \\
        \end{aligned}
    \end{equation*}

    So, the following transformations can be made to the problem

    Which takes us to the \textbf{DARK PART!}\\

    We know that $\frac{\delta Tr[\Lambda G \hat{\Lambda} G^t]}{\delta G} = 2\Lambda G \hat{\Lambda}$\\\\
    We also know that $\frac{\delta Tr(\Lambda^2)}{\delta G} = 0 = \frac{\delta Tr(\hat{\Lambda}^2)}{\delta G}$\\\\
    Finally we know that the answer for a fixed $\hat{\Lambda}$ is $G=I$\\

    \textbf{END OF DARK PART}\\

    This result implies that

    Now, coming back to the original problem this information can be used to
    simplify the objective function

    To minimize the resulting expression, the first $p$ eigenvalues of $\Lambda$
    can be used to define $\hat{\Lambda}$. And finally the solution to the
    problem can be found by the following procedure

    which implies that $Y = \hat{\Lambda}^{1/2} V^t$.

    \vspace{0.25cm}

    This method gives a way to find a matrix $D$ that encodes the distances
    between points that are in a low dimensional space. Now a procedure that
    takes a distance matrix as input and finds a corresponding configuration
    of points is described. The distance matrix is defined as follows

@@

    The unknown of this problem is a matrix $X$ that cotains the position of the
    $n$ points in the $d$ dimensional space.

@@

    But there is a restriction to make sure that the matrix is centered at 0 and
    avoid indeterminations due to the fact that any point can be choosed as center.

@@

    the first step to solve the problem is rewrite each entry of matrix $D$ in
    a vector form

@@

    Then, each component $d_{ij}^2$ can be expanded as follows

@@

    The goal is to express all the dot products of $x$ vectors in terms of the
    $d_{ij}^2$ distances. To achieve this goal, consider the following expression

@@

    But, using the restriction, one term can be canceled

@@

    Finally, the first approximation to the solution of the problem can be
    constructed

@@

    But this expression has a term that can be treated as duplicated, so the
    following propertie can be applied to simplify the expression

    \begin{equation*}
        \begin{aligned}
            \sum_{i=1}^n x_i^t x_i &= \sum_{j=1}^n x_j^t x_j\\
            x_i^t x_j =
            \frac{1}{2} \Bigg [ \frac{1}{n} \sum_{j=1}^n d_{ij}^2 - d_{ij}^2 + &\frac{1}{n} \sum_{i=1}^n d_{ij}^2 - \frac{2}{n} \sum_{i=1}^n x_i^t x_i \Bigg ]\\
        \end{aligned}
    \end{equation*}

@@

    Now the goal is completed, and the dot product between the $X$ components
    can be expressed in terms of distance operations

    \begin{equation*}
        \begin{aligned}
            x_i^t x_j
            &=
            \frac{1}{2} \left[ \frac{1}{n} \sum_{i=1}^n d_{ij}^2 - d_{ij}^2 + \frac{1}{n} \sum_{j=1}^n d_{ij}^2 - \frac{1}{n^2}\sum_{i=1}^n \sum_{j=1}^n d_{ij}^2 \right]\\
        \end{aligned}
    \end{equation*}

    This computation can also be expressed in matrix form to simplify the formula.
    First the following definitions must take place

@@

    Then, a matrix $B$ can be defined as

@@

    This formula gives a way to compute all the dot products between the points
    that makes up the matrix $X$. Because of this, the matrix $B$ can also
    be expressed as

@@

    Which gives a way to compute all the points of the matrix $X$ finding a low
    dimensional representation of the original dataset.

\begin{algorithm}
    $n$ = number of points\\
    $p$ = output dimensions\\
    $k$ = number of neighbors\\
    \vspace{0.5cm}
    $D_e$ = compute\_euclidean\_distance(data)\\
    $G$ = zeros($n$) //graph adjacency matrix\\
    \vspace{0.5cm}
    \For{i in 1:n}{
        $x_i$ = data($i$)\\
        $k_i$ = get\_nearest\_neighbors($x_i$, $k$)\\
        $G(i,k_i)$ = $D_e$($i$, $k_i$)\\
        $G(k_i,i)$ = $D_e$($i$, $k_i$)\\
    }
    \vspace{0.5cm}
    $H$ = $I - \frac{1}{n} e e^t$\\
    $D^x$ = compute\_distances($G$)\\
    $A$ = -$\frac{1}{2}$ pow\_elementwise($D^x$,2)\\
    $B$ = $HAH$\\
    $V$ = get\_eigenvectors($B$)\\
    $\Lambda$ = get\_eigenvalues($B$)\\
    $V$, $\Lambda$ = select\_smallers($p$, $V$, $\Lambda$)\\
    $Y$ = $\Lambda^{1/2} V^t$

    \vspace{0.5cm}

    \caption{Computation of low dimensional representation}
    \label{alg1}
\end{algorithm}

\cite{matlab}

    \section{Apendix}

@@

    Consider a matrix B that is computed as $A^tA$

@@

    Then, an element of B can be computed as

@@

\bibliographystyle{unsrt}
\bibliography{Article_mds}
